---
title: "Selection test"
author: "Lorena Lorenzo Fernández"
date: "25 March, 2021"
output:
csl: 
editor_options:
  chunk_output_type: console
---

## 1. Rehh, an R package for analysing iHS and XP-EHH

### 1.1 Prepare data

Once I have filtered, phased and polarized data (see variant_filtering.md , phasing.md or polarization.md), I can start with selection analyses. The vcf file containing such information is:

`goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf`

The next step is to get a subset data to try in Rstudio the code for the package rehh (see tryals_rehh.R)

Once did it, I have to get a vcf per sp and per chr to process data.

While doing this, I realized that GATK SelectVariants needed the chr length specified in the header so I had to modify the vcf and try again ( I took this info from genomics-a server vcf_filtering). After that, the next problem is de GT info, so I had to make again the header including this info.

``` bash
#Separate per sp vcf
species=(lc ll lp lr)

for sp in ${species[@]}
  do
    #Name per sp samples
    samples=($(grep -m1 "#CHR" /home/llorenzo/goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf | tr '\t' '\n' | grep "${sp}"))
    /opt/gatk-4.1.0.0/gatk SelectVariants \
      -R /GRUPOS/grupolince/reference_genomes/felis_catus_genome/Felis_catus.Felis_catus_9.0.dna.toplevel.fa \
      -V /home/llorenzo/goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf \
      $(for i in ${samples[@]}; do echo "-sn ${i}"; done) \
      -O /home/llorenzo/${sp}_goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf
  done
  
#Separate per chr sp vcf
for sp in ${species[@]}
do
  echo "$sp"
  CHR=($(cat /GRUPOS/grupolince/reference_genomes/felis_catus_genome/Felis_catus.Felis_catus_9.0.dna.toplevel.fa.fai | cut -f 1 | uniq))
  for chr in ${CHR[@]:0:20}
    do
      echo "$chr"
      grep -E "^(#|${chr})" /home/llorenzo/${sp}_goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf \
      > /home/llorenzo/${chr}_${sp}_goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf
      echo "Done $chr for $sp vcf"
    done
done
```

### 1.2 iHS analyses

We will run in parallel the code for calculating ihs in each chr in the genomics-b server (See ihs.R). With this we will get two plots with the ihs results (ihs score and p-value) per chr and sp.

```{bash}
CHR=($(cat /GRUPOS/grupolince/reference_genomes/felis_catus_genome/Felis_catus.Felis_catus_9.0.dna.toplevel.fa.fai | cut -f 1 | uniq))

for chr in ${CHR[@]:0:18}
    do
    echo "$chr"
    screen -dmS "session_$chr" Rscript ./ihs.R $chr
done
```

We also are interested in the genome wide score plots per sp.

```{bash}
species=(lc ll lp lr)

for sp in ${species[@]}
    do
    echo "$sp"
    screen -dmS "session_$sp" Rscript ./ihs_sp.R $sp
    done
```

We found that Kovalaskas et al. 2020 used an adjusted p-value to correct for False Discovery Rate (FDR) using Benjamini-Hochberg method. After investigating about the method and the meaning I tried to integrate in my analyses. The option has to be marked in the ihh2ihs function (See ihs_sp.R):

#### **Test if fixed variants adds noise to iHS calculation**

From the results we can conclude that the results are the same because fixed markers are previously discarded

```{r}
lc_scan_nofixed<- lc_scan %>% filter(FREQ_A != 1 & FREQ_D != 1)

ihs<- ihh2ihs(lc_scan_nofixed, freqbin = 0.05, p.adjust.method = "fdr")
##OUTPUT##
#Discard focal markers with Minor Allele Frequency equal to or below 0.05 .
#506471 markers discarded.
#962534 markers remaining.

ihs2<- ihh2ihs(lc_scan, freqbin = 0.05, p.adjust.method = "fdr")
##OUTPUT##
#Discard focal markers with Minor Allele Frequency equal to or below 0.05 .
#18262514 markers discarded.
#962534 markers remaining.
```

### 1.3 XP-EHH analyses

Discuss here (relying in previous studies) the difference btw this methods, what should we use and why.

Taken from Talenti2017:

The Rsb statistic compares EHH for the same SNP in two different populations and can provide evidence of selection given the presence of high-frequency or fixed alleles in one population but not on the other (Tang et al. 2007). Similarly, the XP-EHH detected selective sweeps in which one allele had undergone strong directional selection in one population while remaining polymorphic in the population as a whole (Sabeti et al. 2007)

We finally decided to use XP-EHH (See xpehh_sp.R)

```{bash}

#Calculating XP-EHH
screen -dmS "session_xpehh" Rscript ./xpehh_sp.R
```

## 2. Lassip, a program for analysing HFS

### 2.1 Installation

I have tried to install lassip in genomics-b server with several errors because of the operation system version:

```{bash}
[llorenzo@genomics_b linux]$ ./lassip
./lassip: /lib64/libstdc++.so.6: version `GLIBCXX_3.4.20' not found (required by ./lassip)
./lassip: /lib64/libstdc++.so.6: version `CXXABI_1.3.8' not found (required by ./lassip)
./lassip: /lib64/libstdc++.so.6: version `CXXABI_1.3.9' not found (required by ./lassip)
./lassip: /lib64/libstdc++.so.6: version `GLIBCXX_3.4.21' not found (required by ./lassip)
```

We had to use conda to install the libraries needed to make it work. Enrico helped me with this part of the code. Each time I want to use lassip I have to go to lassip/lib directory and call the variable with export (not sure what does the command). Once did this, lassip works but I get this error:

*ERROR: Must provide a map from ind to pop with --pop.*

*ERROR: Window size needs to be greater than 0.*

*ERROR: Window step size needs to be greater than 0.*

So we need a pop.txt file and to specify window size and step (101/10).

```{bash}
#make a sp_ind.txt (pop file) needed for lassip
species=(lc ll lp lr)
for sp in ${species[@]}
  do
  sed "s/$/\t${sp}/" <(grep "#CHR" ${sp}_goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf | tr "\t" "\n" | grep "c_")  > ${sp}_ind.txt
  done
```

### 2.2 saltiLASSI analyses

```{bash}
#Create a screen in genomics-b server
screen -S LASSI

#Lassip installation using conda
conda
source activate lassip
cd /home/llorenzo/.conda/envs/lassip/lib
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:/home/tu_usuario/miniconda3/lib/

#try if it works
/home/llorenzo/lassip-master/bin/linux/lassip

#############################LASSIP###############################
##################################################################

#name de variables (sp and chr)  
species=(lc ll lp lr)
CHR=($(grep -v "#" /home/llorenzo/goodsamples_cat_ref.filter8.vcf | cut -f1 | grep -vE "(MT|KZ|AANG)" | uniq)) 

#run the program in a loop per sp and chr
for sp in ${species[@]}
  do
    echo "$sp"
    for chr in ${CHR[@]:0:19} 
      do
        echo "trying salti-lassi in $chr"        
        /home/llorenzo/lassip-master/bin/linux/lassip --vcf /home/llorenzo/rehh_test/chr_ihs/${chr}_${sp}_goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf --calc-spec --hapstats --salti --winsize 101 --winstep 10 --pop /home/llorenzo/rehh_test/${sp}_ind.txt --out /home/llorenzo/rehh_test/${chr}
      done
       /home/llorenzo/lassip-master/bin/linux/lassip --spectra /home/llorenzo/rehh_test/*${sp}.lassip.hap.spectra.gz --salti --out /home/llorenzo/rehh_test/${sp}_salti
  done
  
#lr as the most genetic diverse sp makes lots of windows w/ the parameters especify before, so I try again increasing windows step 
for sp in ${species[@]}
  do
    echo "$sp"
    for chr in ${CHR[@]:0:19} 
      do
        echo "trying salti-lassi in $chr"        
        /home/llorenzo/lassip-master/bin/linux/lassip --vcf /home/llorenzo/rehh_test/chr_ihs/${chr}_${sp}_goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf --calc-spec --hapstats --salti --winsize 101 --winstep 50 --pop /home/llorenzo/rehh_test/${sp}_ind.txt --out /home/llorenzo/rehh_test/${chr}_50wstep
      done
       /home/llorenzo/lassip-master/bin/linux/lassip --spectra /home/llorenzo/rehh_test/*_50wstep.${sp}.lassip.hap.spectra.gz --salti --out /home/llorenzo/rehh_test/${sp}_salti_50wstep
  done
```

#### **Test if fixed variants adds noise to iHS calculation**

As I test before with rehh, I have to test (maybe I could do it previously) if the variants fixed are causing noise in the selection scan. Thus, variants with AF=0.00 and AF=1.00 must be eliminated before the scan (the vast majority).

```{bash}
#name de variables (sp and chr)  
species=(lc ll lp lr)
CHR=($(grep -v "#" /home/llorenzo/goodsamples_cat_ref.filter8.vcf | cut -f1 | grep -vE "(MT|KZ|AANG)" | uniq)) 

for sp in ${species[@]}
  do
    echo "$sp"
    for chr in ${CHR[@]:0:19} 
      do
        echo "eliminate fixed in $chr"     
        grep -vE "(AF=1.00|AF=0.00)" \
        ${chr}_${sp}_goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf \
        > ${chr}_${sp}_goodsamples_filtered_phased_polarized_nofixed_header_cat_ref.vcf
      done
  done  
  
#################################
#######repeat saltiLASSI#########
#################################

#Lassip installation using conda
conda
source activate lassip
cd /home/llorenzo/.conda/envs/lassip/lib
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:/home/tu_usuario/miniconda3/lib/

#name de variables (sp and chr)  
species=(lc ll lp lr)
CHR=($(grep -v "#" /home/llorenzo/goodsamples_cat_ref.filter8.vcf | cut -f1 | grep -vE "(MT|KZ|AANG)" | uniq)) 

#run salti-lassi
for sp in ${species[@]}
  do
    echo "$sp"
    for chr in ${CHR[@]:0:19} 
      do
        echo "trying salti-lassi in $chr"        
        /home/llorenzo/lassip-master/bin/linux/lassip --vcf /home/llorenzo/rehh_test/chr_ihs/${chr}_${sp}_goodsamples_filtered_phased_polarized_nofixed_header_cat_ref.vcf --calc-spec --hapstats --salti --winsize 101 --winstep 10 --pop /home/llorenzo/rehh_test/${sp}_ind.txt --out /home/llorenzo/rehh_test/${chr}_nofixed
      done
       /home/llorenzo/lassip-master/bin/linux/lassip --spectra /home/llorenzo/rehh_test/*_nofixed.${sp}.lassip.hap.spectra.gz --salti --out /home/llorenzo/rehh_test/${sp}_salti_nofixed
  done
  
##PROVED IT DOESNT CHANGE ANYTHING  
```

## 3. Ploting results

For iHS and XP-EHH, the rehh R package has a manhattanplot function that directly plots the results from both analyses. However, for saltiLASSI results I had to plot with ggplot2 and I did the same with iHS for comparative analyses.

### 3.1 saltiLASSI outliers, cutoffs and plots

```{r}
#!/usr/bin/env Rscript

print("Starting plotting LASSI results code")

#Install dependencies

  ##install.packages("dplyr")         ##only once
  ##install.packages("tidyverse")
  ##install.packages("ggplot2")

  library(dplyr)
  library(tidyverse)
  library(ggplot2)

#Set working directory
  setwd("rehh_test")

#Name variables
  species<- c("lc", "ll", "lp", "lr")


for (sp in species)
{
  #Read data
  df<- read.table(paste0(sp, "_salti.lassip.hap.out"), sep="\t", header=T)
  print(paste("Read data for", sp))
  
  #Group by chr
  data_cum <- df %>% 
    group_by(chr) %>% 
    summarise(max_bp = max(pos)) %>%
    mutate(bp_add = lag(cumsum(max_bp), default = 0)) %>% 
    select(chr, bp_add)

  data <- df %>% 
    inner_join(data_cum, by = "chr") %>% 
    mutate(bp_cum = pos + bp_add)

  axis_set <- data %>% 
    group_by(chr) %>% 
    summarize(center = mean(bp_cum))
 
  print("Acumulated by chr")
  
  #Change the "sp_L" column name to another in order to make it easier to manage
  colnames(data)[12]<- "statistic"
  
  print("rename sp_L column to statistic")
  
  #get the 95% and 99% cut-off of the data
  print(paste(sp, round(length(data$statistic) * 0.05), "top 5% values"))
  print(paste(sp, round(length(data$statistic) * 0.01), "top 1% values"))
  
  cut_95 <- sort(data$statistic)[round(length(data$statistic) * 0.95)]
  cut_99 <- sort(data$statistic)[round(length(data$statistic) * 0.99)]
  
  print(paste(sp, cut_95, "5% cut-off" ))
  print(paste(sp, cut_99, "1% cut-off" ))
  
  print("Cut-offs done")
  
  ##########################Plotting results###################################
  
  #Give a name and size to the output plot
  pdf(paste0(sp,"_saltilassi_cutoffs_manhattanplot.pdf"), width=10, height=5)

  
  #Plot with cutoffs
  print(ggplot(data, aes(x = bp_cum, y = statistic, color = as_factor(chr))) +
      geom_point(alpha = 0.5) +
      geom_hline(yintercept= cut_95, color="grey", linetype="dashed") +
      geom_hline(yintercept= cut_99, color="red", linetype="dashed") +
      scale_x_continuous(label = axis_set$chr, breaks = axis_set$center) +
      labs (x="Chromosome", y=expression(Lambda), title=sp ) +
      theme(plot.title = element_text(hjust = 1, vjust = -10), legend.position = "none"))
  
  #Save the pdf results        
  dev.off()
  print("Manhattanplot done and saved")


  #Plot statistic distribution
  pdf(paste0(sp,"_L_statistic_distplot.pdf"), width=10, height=5)
  
  print(ggplot(data, aes(x= statistic)) +
          geom_freqpoly(binwidth=10) + 
          ylim(0, 20000) +
          labs (title=sp))
  
  dev.off()
  
  print("Distributionplot done and saved")
  
  ###########################Extracting outliers####################################
  #Extract a dataframe with the 1% outliers
  outliers<- data %>% filter(statistic >= cut_99) %>% select(c(1,2,3,5,12))
  write.table(outliers, file=(paste0(sp, "_99outliers")), sep="\t", row.names = FALSE, quote = FALSE)
  
  #Save as 0-based bed file
  bed <- outliers %>% mutate(start= start - 1) %>% select(c(1,2,3))
  write.table(bed, file=(paste0(sp, "_99outliers.bed")), sep="\t", col.names= FALSE, row.names = FALSE, quote = FALSE)
  
  print("outliers df saved")
}
```

After this, I want to merge every overlapping outlier window with bedtools merge:

```{bash}
species=(lc ll lp lr)

for sp in ${species[@]}
  do
    echo "$sp"
    bedtools merge -i ${sp}_99outliers.bed -c 1 -o count \
    > ${sp}_99outliers_merged.bed
  done  
```

### 3.2 iHS outliers, cutoffs and plots

```{r}
#!/usr/bin/env Rscript

print("Starting plotting iHS results code")

#Install dependencies

  ##install.packages("dplyr")         ##only once
  ##install.packages("tidyverse")
  ##install.packages("ggplot2")

  library(dplyr)
  library(tidyverse)
  library(ggplot2)

#Name the variables
species<- c("lc", "ll", "lp", "lr")


for (sp in species)
{
  ihs_scan <- read.table(paste0(sp, "_ihs_scan"), sep="\t", header= TRUE)

  #Now, that I have also saltiLASSI results, I am going to plot ihs results in ggplot2
  #Group by chr
  data_cum <- ihs_scan %>% 
    group_by(CHR) %>% 
    summarise(max_bp = max(POSITION)) %>%
    mutate(bp_add = lag(cumsum(max_bp), default = 0)) %>% 
    select(CHR, bp_add)

  data <- ihs_scan %>% 
    inner_join(data_cum, by = "CHR") %>% 
    mutate(bp_cum = POSITION + bp_add) %>%
    na.omit()

  axis_set <- data %>% 
    group_by(CHR) %>% 
    summarize(center = mean(bp_cum))

  #get the 0.1% of data
  cut_99.9 <- quantile(data$LOGPVALUE,probs=0.999,na.rm=T)
  
  print(paste(sp, round(length(data$LOGPVALUE) * 0.001), "top 0.1% values"))
  print(paste(sp, cut_99.9, "0.1% cut-off" ))
  
  print("Cut-offs done")

 ##########################Plotting results###################################
  #Give a name and size to the output plot
  path<- "/home/llorenzo/selection_scan/rehh_test/plots/"
  pdf(paste0(path, sp,"_ihs_cutoffs_manhattanplot.pdf"), width=10, height=5)
  
  print(ggplot(data, aes(x = bp_cum, y = LOGPVALUE, color = as_factor(CHR))) +
          geom_point(alpha = 0.5) +
          geom_hline(yintercept= cut_99.9, color="grey", linetype="dashed") +
          scale_x_continuous(label = axis_set$CHR, breaks = axis_set$center) +
          labs (x="Chromosome", y="p-adjust iHS", title=sp ) +
          theme(plot.title = element_text(hjust = 1, vjust = -10), legend.position = "none"))
  
  dev.off()

  ###########################Extracting outliers####################################
  #Extract a dataframe with the 0.1% outliers
  outliers<- data %>% filter(LOGPVALUE >= cut_99.9) %>% select(c(1,2,3,4))
  write.table(outliers, file=(paste0(sp, "_ihs_99.9outliers")), sep="\t", row.names = FALSE, quote = FALSE)
  
  print("outliers df saved")
  
  #Save as 0-based bed file
  bed <- outliers %>% mutate(start= POSITION - 1) %>% select(c(1,5,2))
  write.table(bed, file=(paste0(sp, "_ihs_99.9outliers.bed")), sep="\t", col.names= FALSE, row.names = FALSE, quote = FALSE)
  
  print("bed df saved")
}  
```

### 3.3 Intersect between the two methods

```{bash}
species=(lc ll lp lr)

for sp in ${species[@]}
  do
    echo "$sp"
    bedtools intersect \
    -a /home/llorenzo/selection_scan/lassi_test/${sp}_99outliers_merged.bed \
    -b /home/llorenzo/selection_scan/rehh_test/${sp}_ihs_99.9outliers.bed \
    -wao \
    > /home/llorenzo/selection_scan/${sp}_outliers_ihslassi_intersect.bed
  done  
```

Once did this I am going to plot shaded areas from LASSI outliers over the iHS plot results

```{r}

#Name the variables
species<- c("lc", "ll", "lp", "lr")


for (sp in species)
{
  ###LASSI###
  #Read the data
  path <- "/home/llorenzo/selection_scan/lassi_test/"
  lassi_outliers <- read.table(paste0(path, sp, "_99outliers_merged.bed"), sep="\t")
  colnames(lassi_outliers) <- c("chr", "start", "end", "n_windows")
    
  #Group by chr
  data_lassi_cum <- lassi_outliers %>% 
    group_by(chr) %>% 
    summarise(max_bp = max(end)) %>%
    mutate(bp_add = lag(cumsum(max_bp), default = 0)) %>% 
    select(chr, bp_add)

  data_lassi <- lassi_outliers %>% 
    inner_join(data_lassi_cum, by = "chr") %>% 
    mutate(bp_start_cum = start + bp_add) %>%
    mutate(bp_end_cum = end + bp_add) %>%
    na.omit()
  
  ###iHS###
  #Read the data
  path <- "/home/llorenzo/selection_scan/rehh_test/"
  ihs_scan <- read.table(paste0(path, sp, "_ihs_scan"), sep="\t", header= TRUE)

  #Group by chr
  data_ihs_cum <- ihs_scan %>% 
    group_by(CHR) %>% 
    summarise(max_bp = max(POSITION)) %>%
    mutate(bp_add = lag(cumsum(max_bp), default = 0)) %>% 
    select(CHR, bp_add)

  data_ihs <- ihs_scan %>% 
    inner_join(data_ihs_cum, by = "CHR") %>% 
    mutate(bp_cum = POSITION + bp_add) %>%
    na.omit()

  ihs_axis_set <- data_ihs %>% 
    group_by(CHR) %>% 
    summarize(center = mean(bp_cum))

  #get the 0.1% of data
  cut_99.9 <- quantile(data_ihs$LOGPVALUE,probs=0.999,na.rm=T)
  
  print(paste(sp, round(length(data_ihs$LOGPVALUE) * 0.001), "top 0.1% values"))
  print(paste(sp, cut_99.9, "0.1% cut-off" ))
  
  print("Cut-offs done")
  
#Plot ihs with lassi windows  
path<- "/home/llorenzo/selection_scan/"
pdf(paste0(path, sp, "_ihs_lassi_cutoffs_manhattanplot.pdf"), width=10, height=5)

print(ggplot() +
        geom_point(data_ihs, mapping=aes(x = bp_cum, y = LOGPVALUE, color = as_factor(CHR)), alpha = 0.5) +
        geom_hline(yintercept= cut_99.9, color="grey", linetype="dashed") +
        geom_rect(data_lassi, mapping=aes(xmin = bp_start_cum, xmax = bp_end_cum, ymin = -Inf, ymax = Inf), alpha = 0.4, color = as_factor(CHR)) +
        scale_x_continuous(label = axis_set$CHR, breaks = axis_set$center) +
        labs (x="Chromosome", y="p-adjust iHS", title=sp ) +
        theme(plot.title = element_text(hjust = 1, vjust = -10), legend.position = "none"))

dev.off()

print("Plot done and saved")
}
```

Any conclusion can be extracted from this plot. I represented all SNPs from iHS and windows can rarely be seen as they are really small in contrast with the chr length, so as a result, we can only see a vertical line representing the window and lots of points representing SNPs.

So, the next thing I am going to try is to plot per chr outliers windows from saltilassi and the outliers SNPs from iHS.

```{r}
#Name the variables
species<- c("lc", "ll", "lp", "lr")


for (sp in species)
{
  ###LASSI###
  #Read the data
  path <- "/home/llorenzo/selection_scan/lassi_test/"
  lassi_outliers <- read.table(paste0(path, sp, "_99outliers_merged.bed"), sep="\t") 
  colnames(lassi_outliers) <- c("CHR", "START", "END", "N_WINDOWS")
  lassi_outliers <-  filter(lassi_outliers, CHR != "X")
  print("Reading lassi results done")
  
  ###iHS###
  #Read the data
  path <- "/home/llorenzo/selection_scan/rehh_test/"
  ihs_outliers <- read.table(paste0(path, sp, "_ihs_99.9outliers"), sep="\t", header= TRUE)
  colnames(ihs_outliers) <- c("CHR", "POS", "IHS", "PVAL")
  print("Reading iHS results done")
  
  #Plot ihs with lassi windows  
#Define chromosome variable
chromosomes <- c("A1", "A2", "A3", "B1", "B2", "B3", "B4", "C1", "C2", "D1", "D2", "D3", "D4", "E1", "E2", "E3", "F1", "F2")

#loop per chr plot
for (chr in chromosomes)
{
chr_data_ihs <- ihs_outliers %>%
  filter(CHR== chr)
chr_data_lassi <- lassi_outliers %>%
  filter(CHR== chr)
    
path<- "/home/llorenzo/selection_scan/"
pdf(paste0(path, chr, "_", sp, "_ihs_lassi_manhattanplot.pdf"), width=10, height=5)

print(ggplot() +
        geom_point(chr_data_ihs, mapping=aes(x = POS, y = PVAL, alpha = 0.5)) +
        geom_rect(chr_data_lassi, mapping=aes(xmin = START, xmax = END, ymin = -Inf, ymax = Inf), alpha = 0.4) +
        labs (x="Chromosome", y="p-adjust iHS", title=sp ) +
        theme(plot.title = element_text(hjust = 1, vjust = -10), legend.position = "none"))

dev.off()

print("Plot done and saved")
}
}
```

By doing this I discovered that 0.1% of SNPS and 1% of windows is very restrictive. I am going to try with 1% of SNPS and 5% of windows. Well maybe is only in lc

PROBLEM FOUND: Lp has aprx. 8.2M variants, which is much more than expected... I went back to the sp_filter8.vcf and discover that by this point lp has 1.2M variants which is far more feasible. Now I have to traceback where I did it wrong (but unluckily this week I don't have access to CESGA, due to CSIC-cibersecurity problems)

### 3.3.1 PCA before and after phasing

```{bash}
###PCA
#Repeat plink per sp.
species=(lc ll lp lr)

for i in ${species[@]}
  do
    VCF=/home/llorenzo/selection_scan/rehh_test/${i}_goodsamples_filtered_phased_polarized_ACfiltered_header_cat_ref.vcf
    # perform linkage pruning - i.e. identify prune sites
    plink_1.9 --vcf $VCF --double-id --allow-extra-chr \
    --set-missing-var-ids @:# \
    --indep-pairwise 50 10 0.1 --out ${i}_goodsamples_filtered_phased_polarized_cat_ref
    #run PCA in the subset
    plink_1.9 --vcf $VCF --double-id --allow-extra-chr --set-missing-var-ids @:# \
    --extract ${i}_goodsamples_filtered_phased_polarized_cat_ref.prune.in \
    --make-bed --pca --out ${i}_goodsamples_filtered_phased_polarized_cat_ref
  done


```

```{r}
library(tidyverse)
library(viridis)


species<- c("lc", "ll", "lp", "lr")
path <- "~/"

##PREPHASED PCAs##
for (sp in species)
{
  #Read pre-phased data
  # import eigen vec and val
pca <- read_table(paste0(path, sp, "_goodsamples_cat_ref.eigenvec"),
                   col_names = FALSE)
eigenval <- scan(paste0(path, sp, "_goodsamples_cat_ref.eigenval"))  

# remove nuisance column
pca <- pca[,-1]
# set names
names(pca)[1] <- "ind"
names(pca)[2:ncol(pca)] <- paste0("PC", 1:(ncol(pca)-1))

print("named cols")

# first convert to percentage variance explained
pve <- data.frame(PC = 1:(ncol(pca)-1), pve = eigenval/sum(eigenval)*100)
# then make a plot
pdf(paste0(path, sp, "prephased_pve.pdf"))
print(ggplot(pve, aes(PC, pve)) +
        geom_bar(stat = "identity") +
        ylab("Percentage variance explained") + 
        theme_light())
dev.off()
print("pve plot done")

# plot PC1 PC2
pdf(paste0(path, sp, "prephased_pca.pdf"))

print(ggplot() +
   geom_point(pca, mapping=aes(x= PC1, y= PC2, color=ind) , size=4, alpha = 0.5) +
    geom_text(pca, mapping=aes(x= PC1, y= PC2, label=ind), size=3, check_overlap = TRUE) +
    coord_equal() +
    theme() +
    xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) +
    ylab(paste0("PC2 (", signif(pve$pve[2], 3), "%)")))
dev.off()
print("pca plot done")
}

##PHASED PCAs##
for (sp in species)
{
#Read phased data
pca <- read_table(paste0(path, sp, "_goodsamples_filtered_phased_polarized_cat_ref.eigenvec"),
                   col_names = FALSE)
eigenval <- scan(paste0(path, sp, "_goodsamples_filtered_phased_polarized_cat_ref.eigenval"))

print("read data")
# remove nuisance column
pca <- pca[,-1]
# set names
names(pca)[1] <- "ind"
names(pca)[2:ncol(pca)] <- paste0("PC", 1:(ncol(pca)-1))

print("named cols")

# first convert to percentage variance explained
pve <- data.frame(PC = 1:(ncol(pca)-1), pve = eigenval/sum(eigenval)*100)
# then make a plot
pdf(paste0(path, sp, "phased_pve.pdf"))
print(ggplot(pve, aes(PC, pve)) +
        geom_bar(stat = "identity") +
        ylab("Percentage variance explained") + 
        theme_light())
dev.off()
print("pve plot done")

# plot PC1 PC2
pdf(paste0(path, sp, "phased_pca.pdf"))

print(ggplot() +
   geom_point(pca, mapping=aes(x= PC1, y= PC2, color=ind) , size=4, alpha = 0.5) +
    geom_text(pca, mapping=aes(x= PC1, y= PC2, label=ind), size=3, check_overlap = TRUE) +
    coord_equal() +
    theme() +
    xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) +
    ylab(paste0("PC2 (", signif(pve$pve[2], 3), "%)")))
dev.off()
print("pca plot done")
}
```

```{bash}
bcftools view -e 'INFO/AF=1.00 | INFO/AF=0.00' *_lp_whatshap_phased_duplicated.vcf | wc -l
```

```{bash}

#LOOKING FOR THE ERROR

species=(lc lp lr)
CHR=($(cat $STORE2/reference_genomes/Felis_catus_Ref/Felis_catus.Felis_catus_9.0.dna.toplevel.fa.fai | cut -f 1 | uniq))

for sp in ${species[@]}
  do
    for chr in ${CHR[@]:0:19} 
        do
          grep -v "#" ${chr}_${sp}_whatshap_phased_duplicated.vcf  >> ${sp}_tryal_nvariants.vcf
      done
      wc -l ${sp}_tryal_nvariants.vcf
      cat <(grep "#" A1_${sp}_whatshap_phased_duplicated.vcf) ${sp}_tryal_nvariants.vcf > ${sp}_tryal_nvariants_header.vcf    
      bcftools view -e 'INFO/AF=1.00 | INFO/AF=0.00' ${sp}_tryal_nvariants_header.vcf | wc -l
    done
    

for sp in ${species[@]}
  do
    for chr in ${CHR[@]:0:19} 
        do
          grep -v "#" ${chr}_${sp}_shapeit_phased_duplicated_cat_ref.vcf  >> ${sp}_tryal_shapeit_nvariants.vcf
      done
      wc -l ${sp}_tryal_shapeit_nvariants.vcf
      cat <(grep "#" $LUSTRE/phasing/intermediate_files/A1_${sp}_whatshap_phased_duplicated.vcf) ${sp}_tryal_shapeit_nvariants.vcf > ${sp}_tryal_shapeit_nvariants_header.vcf    
      bcftools view -e 'INFO/AF=1.00 | INFO/AF=0.00' ${sp}_tryal_shapeit_nvariants_header.vcf | wc -l
    done  
   
```
